Simple repression draft

### 2020-05-26 doc/appendix_03_regulated_promoters.tex
update minor things such as figure naming
making lots of progress on the bursty promoter derivation

% !TEX root = ./busty_transcription.tex
\section{Bursty promoter models - generating function solutions and numerics}
\label{sec:gen_fcn_appdx}

@@

We would like to derive the full probability distribution of
mRNA copy number per cell for a model combining simple repression
with a promoter featuring bursty transcription initiation.
Before tackling mRNA and repression together, let us
just consider the bursty consititutive promoter,
model 5 in~\fig{fig:constit_cartoons}
\mmnote{replicating Charlotte's notes; she says it's worked out
in~\cite{Paulsson2000}, but I don't see it, at least not in notation I can
comprehend. Need to find a cite for this,
or maybe just cite Charlotte, personal comm, since that's where I actually got it\dots}.
The master equation of interest is

@@

\deriv{t}p(m,t) = (m+1)\gamma p(m+1,t) - m\gamma p(m,t) - r p(m,t)
        + r \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),
\label{eq:1state_unreg_003}

@@

where $G_{k}(\theta)$ is the geometric distribution defined as

@@

G_{k}(\theta) = \theta(1 - \theta)^k, \, k\in\{0,1,2,\dots\}.

@@

With this convention, the mean burst size $b = (1-\theta)/\theta$.
$\gamma$ and $r$ are mRNA degradation rates and transcription burst
initiation rates, resp. The last term represents all ways the system could
end up with $m$ mRNAs, having started the burst with $m^\prime$. Define
$\lambda = r/\gamma$ and nondimensionalize time by $\gamma$, giving

@@

\deriv{t}p(m,t) = (m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
        + \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t),

@@

The probability generating function is defined as

@@

F(z,t) = \sum_{m=0}^\infty z^m p(m,t).

@@

Multiply both sides of the CME by $z^m$ and sum over all $m$ to get

@@

\pderiv[F]{t} = \sum_{m=0}^\infty z^m
(m+1)p(m+1,t) - m p(m,t) - \lambda p(m,t)
+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime}(\theta) p(m^\prime,t)

@@

The first two terms can be rewritten as derivatives of the
generating function, i.e.,

@@

\sum_{m=0}^\infty z^m (m+1)p(m+1,t)
\pderiv{z} \sum_{m=0}^\infty z^{m+1} p(m+1,t)
\pderiv[F]{z}

@@

and
\sum_{m=0}^\infty z^m mp(m,t)
z\pderiv{z} \sum_{m=0}^\infty z^m p(m,t)
z\pderiv[F]{z}.
The final term is the most trouble. The trick is to reverse
the default order of the sums, as in
&\lambda \sum_{m=0}^\infty \sum_{m^\prime=0}^m
        z^m \theta(1 - \theta)^{m-m^\prime} p(m^\prime,t)
\\=
&\lambda\theta \sum_{m^\prime=0}^\infty z^{m^\prime} p(m^\prime,t)
        \sum_{m=m^\prime}^\infty
        z^{m-m^\prime} (1 - \theta)^{m-m^\prime}
\\=
&\frac{\lambda\theta F}{1 - (1 - \theta)z}.
This works because after reordering the sums, the sum on $m$ is just
a geometric series. Its result is independent of $m^\prime$, so the sum on
$m^\prime$ reduces to the definition of the generating function.

>>>>

% !TEX root = ./bursty_transcription.tex
\section{Bursty promoter models - generating function solutions and numerics}
\label{sec:gen_fcn_appdx}

@@

The objective of this section is to write down the steady-state mRNA
distribution for model 5 in Figure~\ref{fig2:constit_cartoons}. Our claim is
that this model is rich enough that it can capture the expression pattern of
bacteria unregulated promoters regulated by the $\sigma_{70}$ sigma factor.
Figure~\ref{figS1:bursty_one_state} shows two different schematic
representations of the model. Figure~\ref{figS1:bursty_one_state}(A) shows the
promoter cartoon model with burst initiation rate $k_i$, mRNA degradation rate
$\gamma$, and mean burst size $b$. For our derivation of the chemical master
equation we will focus more on Figure~\ref{figS1:bursty_one_state}(B). This
representation is intended to highlight that bursty gene expression allows
transitions between mRNA count $m$ and $m'$ even with $m - m' > 1$.

\begin{figure}[h!]
\centering
\includegraphics{../figures/si/figS0X_bursty_states.pdf}
\caption{
\textbf{Bursty transcription for unregulated promoter.}
(A) Schematic of the one-state bursty transcription model. Rate $k_i$ is the
bursty initiation rate, $\gamma$ is the mRNA degradation rate, and $b$ is the
mean burst size. (B) Schematic depiction of the mRNA count state transitions.
The model in (A) allows for transitions of $> 1$ mRNA counts with probability
$G_{m-m'}$.}
\label{figS1:bursty_one_state}
\end{figure}

To derive the master equation we begin by thinking the possible state
transitions to ``enter'' state $m$. There are two possible paths to jump from an
mRNA count $m' \neq m$ to a state $m$ in a small time window $\Delta t$:
\begin{enumerate}
        \item By degradation of a single mRNA, jumping from $m+1$ to $m$.
        \item By producing $m-m'$ mRNA for $m' \in \{0, 1, \ldots, m-1\}$.
\end{enumerate}
For the ``exit'' states from $m$ into $m' \neq m$ during a small time window
$\Delta t$ we also have two possibilities:
\begin{enumerate}
        \item By degradation of a single mRNA, jumping from $m$ to $m-1$.
        \item By producing $m'-m$ mRNA for $m'-m \in \{1, 2, \ldots\}$.
\end{enumerate}
This implies that the probability of having $m$ mRNA at time $t + \Delta t$ can
be written as
\begin{equation}
\begin{split}
p(m, t + \Delta t) = &p(m, t)
+ \overbrace{\gamma \Delta t (m + 1) p(m + 1, t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma \Delta t m p(m, t)}^{m \rightarrow m - 1} \\
&+ \overbrace{k_i \Delta t \sum_{m'=0}^{m-1} G_{m-m'} p(m', t)}^
{m'\in \{0, 1, \ldots m-1\} \rightarrow m}
- \overbrace{k_i \Delta t \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t)}^
{m \rightarrow m'\in \{m+1, m+2, \ldots\}},
\end{split}
\label{eq:si_master_deltat}
\end{equation}
where we indicate $G_{m'-m}$ as the probability of having a burst of size
$m'-m$. We suggestively use the letter $G$ as we will assume that these bursts
sizes are geometrically distributed with parameter $\theta$. This is written as
\begin{equation}
G_{k} = \theta (1 - \theta)^k\; \text{for } k \in \{0, 1, 2, \ldots \}.
\end{equation}
What this implies is that for a geometrically distributed burst size we have a
mean burst size $b$ of the form
\begin{equation}
b \equiv \left\langle m' - m \right\rangle 
= \sum_{k=0}^\infty k \theta (1 - \theta)^k = {1 - \theta \over \theta}.
\end{equation}

To clean up Equation~\ref{eq:si_master_deltat} we can send the first term on the
right hand side to the left, and divide both sides by $\Delta t$. Upon taking
the limit where $\Delta t \rightarrow 0$ we can write
\begin{equation}
{d \over dt}p(m, t) = (m + 1) \gamma p(m + 1, t)
- m \gamma p(m, t)
+ k_i \sum_{m'=0}^{m-1} G_{m-m'} p(m', t) 
- k_i \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, t).
\end{equation}
Furthermore, given that the timescale for this equation is set by the mRNA
degradation rate $\gamma$ we can divide both sides by this rate, obtaining
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau),
\label{eq:si_master_ode}
\end{equation}
where we defined $\tau \equiv t \times \gamma$, and $\lambda \equiv k_i/\gamma$.
The last term in Eq.~\ref{eq:si_master_ode} sums all burst sizes except for a 
burst of size zero. We can re-index the sum to include this term, obtaining
\begin{equation}
\lambda \sum_{m'=m + 1}^{\infty} G_{m'-m} p(m, \tau) = \lambda p(m, t) \left[
\underbrace{\sum_{m'=\textcolor{red}{m}}^{\infty}G_{m'-m}}
_{\text{re-index sum to include burst size zero}} -
\underbrace{G_0}_{\text{subtract extra added term}}\right].
\end{equation}
Given the normalization of the geometric distribution, adding the probability of
all possible burst sizes -- including size zero since we re-indexed the sum --
allows us to write
\begin{equation}
\sum_{m'=m}^{\infty}G_{m'-m} - G_0 = 1 - G_0.
\end{equation}
Substituting this into Eq.~\ref{eq:si_master_ode} results in
\begin{equation}
{d \over d\tau}p(m, \tau) = (m + 1) p(m + 1, \tau)
- m p(m, \tau)
+ \lambda \sum_{m'=0}^{m-1} G_{m-m'} p(m', \tau) 
- \lambda p(m, \tau) \left[ 1 - G_0 \right].
\label{eq:si_master_ode_2}
\end{equation}
To finally get at a more compact version of the equation notice that the third
term in Eq.~\ref{eq:si_master_ode_2} includes burst from size $m'-m = 1$ to size
$m' - m = m$. We can include the term $p(m, t) G_0$ in the sum which allows
bursts of size $m' - m = 0$. This results in our final form for the chemical
master equation
\begin{align}
{d \over d\tau}p(m, \tau) = 
(m + 1) p(m+1, \tau)
- m p(m, \tau) - 
\lambda p(m, \tau)
+ \lambda \sum_{m^\prime=0}^m G_{m-m^\prime} p(m^\prime, \tau).
\label{eq:si_master_unreg}
\end{align}

In order to solve Eq.~\ref{eq:si_master_unreg} we will use the generating
function method \mrm{reference?}. The probability generating function is defined
as

@@

F(z,t) = \sum_{m=0}^\infty z^m p(m,t),

@@

where $z$ is just a dummy variable that will help us later on to obtain the
moments of the distribution. Applying $F(z, t)$ on both sides of 
Eq.~\ref{eq:si_master_unreg} results in
\sum_m z^m {d \over d\tau} p(m, \tau) = 
\sum_m z^m \left[ 
- m p(m, \tau) 
+ (m + 1) p(m + 1, \tau) 
+ \lambda \sum_{m' = 0}^m G_{m-m'} p(m', \tau) - \lambda p(m, \tau)
where we use $\sum_m \equiv \sum_{m=0}^\infty$. We can distribute the sum and
use the definition of $F(z, t)$ to obtain
{d F(z, \tau) \over d\tau} =
- \sum_m z^m m p(m, \tau) +
+ \sum_m z^m (m + 1) p(m + 1, \tau)
+ \lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau)
- \lambda F(z, \tau).
\label{eq:si_generating_01}
We can make use of properties of the generating function to write everything in
terms of $F(z, \tau)$: the first term on the right hand side of
Eq.~\ref{eq:si_generating_01} can be rewritten as

@@

\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=
\sum_{m} z \frac{\partial z^{m}}{\partial z} p(m, \tau), \\
&=\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
&=z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
&= z {\partial F(z, \tau) \over \partial z}.

@@

For the second term on the right hand side of Eq.~\ref{eq:si_generating_01} we
define $k \equiv m + 1$. This allows us to write

@@

\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=
\sum_{k=1}^{\infty} z^{k-1} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
&=\frac{\partial F(z)}{\partial z}.

@@

The third term in Eq.~\ref{eq:si_generating_01} is the most trouble. The trick
is to reverse the default order of the sums as
\sum_{m=0}^{\infty} \sum_{m'=0}^{m} = \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty}.
This results in

@@

\lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau) =
\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau),

@@

where we also substituted the definition of the geometric distribution $G_{k} =
\theta (1 - \theta)^k$. Redistributing the sums we can write

@@

\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau) = 
\lambda \theta \sum_{n=0}^{\infty}(1-\theta)^{m'} P(m', \tau) 
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m}.
\label{eq:si_generating_02}

@@

We know that $1 - \theta \in (0, 1)$, and that $z$ is a dummy variable that we
usually evaluate as $z=1$. With these two facts we can safely state that
$| z (1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in 
Eq.~\ref{eq:si_generating_02} as

@@

\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &= 
\sum_{n=0}^{\infty} \left[z (1-\theta)\right]^{n + m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\left( {1 \over 1 - z (1 - \theta)} \right),

@@

where we use the geometric series since, as stated before,$| z (1 - \theta) | <
1$
